# Chapter 2 - Loading, Exploring and Modeling the Learning2014 Dataset
##### This week's object of study was the Learning2014 dataset. The Learning2014 dataset contains the results of a survey whose aim was to map out how different things affect the learning of university students. The data was collected from the students of Introduction to Social Statistics in the fall of 2014 (a course taught by Kimmo Vehkalahti in the University of Helsinki). I will analyse the Learning2014 data by going trough the typical following phases.

* ***Load the data***
* ***Explore the data***
* ***Model the data***

***


```{r, echo = FALSE}
### Install the packages which are needed
### install.packages("ggplot2")
library(ggplot2)

### install.packages("GGally")
library(GGally)
```

```{r, echo = FALSE}
### Load in the dplyr library that has many usefule functions for data wrangling
library(dplyr)
```
## 1. Load the data
1. Read the data(-frame) from a .csv-file with the read.table -function

```{r, echo = FALSE}

lrn2014 <- read.table("data/learning2014.csv", sep=",",  header=TRUE)

```

```{r, echo=FALSE}
###lrn2014 <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", sep=",", header=TRUE)
```

## 2. Exploring and Visualizing the data
1. Show the structure of the dataframe with the str() function 

```{r}
str(lrn2014[-1])



```
2. Show the dimensions of the dataframe with the dim() function
```{r}
dim(lrn2014[-1])
```
3. Show 10 first observations by using the head() -function
```{r}
head(lrn2014[-1], 10)
```
#### Visualize the data by plotting it
##### Show a graphical overview and summaries of the data (2.2)

```{r, echo= FALSE}
#ggplot(lrn2014, aes(x= gender, y=points)) + geom_point()
#ggplot(lrn2014, aes(x= age, y=points)) + geom_point()
#ggplot(lrn2014, aes(x= attitude, y=points)) + geom_point()
#ggplot(lrn2014, aes(x= deep, y=points)) + geom_point()
#ggplot(lrn2014, aes(x= stra, y=points)) + geom_point()
#ggplot(lrn2014, aes(x= surf, y=points)) + geom_point()

```

```{r}
ggpairs(lrn2014[-1], mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

### p2 <- pairs(learning2014, , col=learning2014$gender) 
```
## 3. Modeling the Data
#### Build a data model

#### Exam points as the dependent variable and age, gender and attitude as explanatory variables

```{r}

ggplot(lrn2014, aes(x= Age, y=Points, col=attitude)) + geom_point() + facet_grid(. ~ gender) + ggtitle("Age, Attitude and Gender versus Course Points")
```


##### Fitting the regession model to the graphic shown above

```{r}

ggplot(lrn2014, aes(x= Age, y=Points, col=attitude)) + geom_point() + facet_grid(. ~ gender) + stat_smooth(method="lm", col="red") + ggtitle("Age, Attitude and Gender versus Course Points")
```

##### Adding a smooth line to the graphic shown above

```{r}

ggplot(lrn2014, aes(x= Age, y=Points, col=attitude)) + geom_point() + geom_smooth(method="lm")  + ggtitle("Age, Attitude and Gender versus Course Points") + facet_grid(. ~ gender)
```


## 4. Summarizing the Fitted Model and Interpreting the Model Parameters

#### The regression model of Age, Attitude and Gender versus Course Points with the summary of the fitted model
From the summary of the regression model below it can be deduced that age and male gender are  negatively correlated to course points while attitude affects course points in a very positive way.
```{r}
my_model2 <- lm(Points ~ Age + gender + attitude, data = lrn2014)
summary(my_model2)
```
  

#### Diagnostic plots
He are some diagnostic plots visualizing the model parameters: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage.  
```{r, echo=FALSE}
my_model2 <- lm(Points ~ Age + gender + attitude, data = lrn2014)
par(mfrow = c(2,2))
plot(my_model2, which = c(1:2, 5),  caption = list("Residuals vs Fitted", "Normal Q-Q",  "Residuals vs Leverage"))
```

